The Three-HITs Theorem

### 2017-03-27 Paper/Paper.tex
Include definitions of P[A], P[f], and lifting.
It is bad manners to refer the reader to some other paper for something
that is constantly used everywhere in the paper. So I included these.

@@

\newcommand{\idmap}[0]{\Boperator{Id}}
\newcommand{\refl}[0]{\Boperator{refl}}

@@

\newcommand{\hpushout}[0]{\Boperator{hpushout}}
\newcommand{\hocolim}[0]{\Boperator{hocolim}}
\newcommand{\seg}[0]{\Boperator{seg}}
\newcommand{\inl}[0]{\Boperator{inl}}
\newcommand{\inr}[0]{\Boperator{inr}}
\newcommand{\glue}[0]{\Boperator{glue}}
\newcommand{\inc}[0]{\Boperator{inc}}
\newcommand{\com}[0]{\Boperator{com}}
\newcommand{\Type}[0]{\operatorname{\textsc{Type}}}
\newcommand{\pt}[0]{*}

@@

\section{Preliminaries}
\label{sec:prelims}
Let us briefly recall the scheme of higher inductive types which we shall use \cite{basoldhigher}.
First of all, we need \emph{polynomial functors}.
\begin{definition}
Let $X$ be a variable.
Then a \emph{polynomial} is given by the following grammar.
\[
F, G ::= B \ | \ X \ | \ F \times G \ | \ F + G
\]
where $B : \Type$.
\end{definition}
A possible extension would be to use arbitrary containers as in \cite{abbott2005containers}, but we shall refrain to do so.
With that extension the given proof requires the axiom of choice which generally does not hold in type theory.
One can prove that polynomials are functors, and that given a type family $Y : T \rightarrow \Type$ and a polynomial $F$, we get a lift $\bar{F} : F \> T \rightarrow \Type$.
We will write $F \> f$ to denote the application of a polynomial $F$ to a map $f$.
Also, given a dependent function $f : \prod (x : T), Y \> x$, one can define $\bar{F} \> f : \prod (x : H \> T), \bar{H} \> Y \> x$.
Precise definitions of these are given in \cite{basoldhigher}.

@@

\section{Conclusion and Further Work}
Higher inductive types can thus be constructed if we have the interval, homotopy pushout and homotopy colimits.
Since the construction is done in homotopy type theory without assuming extra axioms, the proof is constructive.
Notice that the only inductive types we needed were the sum and product types rather than the general class of W-types or those described inductive schemes.
Hence, we can also deduce the existence of all inductive types from this theorem if we just have sums, products, and the three HITs.

>>>>

% In order to avoid macro pollution, we should comment each macro
% so that we know what it is about

\newcommand{\tyFont}[1]{\textsf{#1}} % the font used to display types
\newcommand{\termFont}[1]{\textsf{#1}} % the font used to term constants
\newcommand{\nat}{\tyFont{N}} % the type of natural numbers
\newcommand{\lift}[1]{\overline{#1}} % the lifting of a type family by a polynomial
\newcommand{\defeq}{\mathrel{{:}{\equiv}}} % definitional equality
\newcommand{\idmap}[1]{\termFont{id}_{#1}} % identity map
\newcommand{\inl}[1]{\termFont{inl}\,#1} % left inclusion into a sum
\newcommand{\inr}[1]{\termFont{inr}\,#1} % right inclusion into a sum

\newcommand{\dprod}[1]{{\textstyle\prod_{#1}}\,} % dependent product
\newcommand{\dsum}[1]{{\textstyle\sum_{#1}}\,} % dependent sum

% Below are macros which have not been documented and are possibly
% unused (probably not)

@@

\newcommand{\refl}[0]{\Boperator{refl}}

@@

\newcommand{\hpushout}[0]{\Boperator{hpushout}}
\newcommand{\hocolim}[0]{\Boperator{hocolim}}
\newcommand{\seg}[0]{\Boperator{seg}}
\newcommand{\glue}[0]{\Boperator{glue}}
\newcommand{\inc}[0]{\Boperator{inc}}
\newcommand{\com}[0]{\Boperator{com}}
\newcommand{\Type}[0]{\operatorname{\textsc{Type}}}
\newcommand{\pt}[0]{*}

@@

\section{A definition of HITs}
\label{sec:prelims}

In this section we provide the definitions of higher inductive types that we shall
consider. We recall the scheme for specifying higher inductive types
from~\cite{basoldhigher}.

\begin{definition}
  A \emph{polynomial (expression)} in a type variable $X$ is a built inductively from the
  variable $X$, type constants~$C$, binary products $\times$, and binary sums $+$.
\end{definition}

For example, if $\nat$ is the type of natural numbers, the expression
$(X + \nat) \times X$ is a polynomial expression in~$X$. We write $P[X]$ to indicate a
polynomial in variable $X$. Given a type $A$, we may substutute $A$ for $X$ to obtain a
type $P[A]$, as follows:
%
\begin{align*}
  X[A] &\defeq A, &
  (P_1 \times P_2)[A] &\defeq P_1[A] \times P_2[A], \\
  C[A] &\defeq C, &
  (P_1 + P_2)[A] &\defeq P_1[A] + P_2[A].
\end{align*}
%
In a similar fashion a map $f \> A \to B$ is transformed to a map $P[f] : P[A] \to P[B]$:
%
\begin{align*}
  X[f] &\defeq f, &
  (P_1 \times P_2)[f] \, (x,y) &\defeq (P_1[f]\, x, P_2[f] \, y), \\
  C[f] &\defeq \idmap{C}, &
  (P_1 + P_2)[f](\inl{x}) &\defeq \inl{(P_1[f]\,x)}, \\
  & &
  (P_1 + P_2)[f](\inr{y}) &\defeq \inr{(P_2[f]\,y)}.
\end{align*}
%
Somewhat less obviously, a polynomial acts on a type family $B : A \to \Type$ to
give its lifting $\lift{P}[B] : P[A] \to \Type$,
%
\begin{align*}
  \lift{X}[B] \, u &\defeq B\,u, &
  \lift{P_1 \times P_2}[B] \, u \defeq (\lift{P_1}[B]\,u) \times (\lift{P_2}[B]\,u), \\
  \lift{C}[B] \, u &\defeq C, &
  \lift{P_1 + P_2}[B] \, (\inl{u})\defeq \lift{P_1}[B]\,u, \\
  & &
  \lift{P_1 + P_2}[B] \, (\inr{u})\defeq \lift{P_2}[B]\,u,
\end{align*}
%
while a map $f \> \dprod{x \> A} B_1\,x \to B_2\,x$ lifts to
$\lift{P}[f] \> \dprod{u : P[A]} \lift{P}[B_1]\,u \to \lift{P}[B_2]\,u$:
%
\begin{align*}
  \lift{X}[f] \, u &\defeq f \, u &
  \lift{P_1 \times P_2}[f] \, (u_1, u_2) &\defeq (\lift{P_1}[f]\,u_1, \lift{P_2}[f]\,u_2) \\
  \lift{C}[f] \, u &\defeq u &
  \lift{P_1 + P_2}[f] \, (\inl{u})\defeq \lift{P_1}[f]\,u, \\
  & &
  \lift{P_1 + P_2}[f] \, (\inr{u})\defeq \lift{P_2}[f]\,u.
\end{align*}
%
There is of course a strong analogy between the action of $P[X]$ and a polynomial functor,
but we hesitate to call $P[X]$ a functor as there is no category to speak of, and in any
case the manipulations we have just described are syntactic.

@@

\section{Conclusion and Further Work}

MOVED FROM BEGINNING: A possible extension would be to use arbitrary containers as in \cite{abbott2005containers}, but we shall refrain to do so.
With that extension the given proof requires the axiom of choice which generally does not hold in type theory.

Higher inductive types can thus be constructed if we have the interval, homotopy pushout and homotopy colimits.
Since the construction is done in homotopy type theory without assuming extra axioms, the proof is constructive.
Notice that the only inductive types we needed were the sum and product types rather than the general class of W-types or those described inductive schemes.
Hence, we can also deduce the existence of all inductive types from this theorem if we just have sums, products, and the three HITs.
